---
title: "Does Open Data Exist on Somerville's Concrete-Block Garages?"
format: html
author: corey
---

I started looking for data on Somerville's concrete-block garages at the Somerville Historical Commission.
These structures define Somerville for me because they seem to stick out behind or next to or beneath every house in the city.
The Historical Commission assured me that the city didn't have any data on the detached garages.
Instead, they recommended I look at the Library of Congress' archive of Sanborn Fire Insurance maps.

Before I did that, however, I thought I'd look into the open data that the city publishes:
* buildings
* tax parcels
* impervious surfaces

## Buildings

```{r}
library(tidyverse)
library(sf)
library(gt)
```

```{r}
temp <- tempfile()
tdir <- tempdir()
download.file("https://data.somervillema.gov/download/uzdd-gyjv/application%2Fx-zip-compressed", temp)
unzip(temp, exdir = tdir)
buildings <- st_read(paste0(tdir, "/Buildings.shp"))
```

```{r}
head(buildings)
```

```{r}
plot(st_geometry(buildings[c(1:50),]))
```

## Tax parcels

```{r}
temp <- tempfile()
tdir <- tempdir()
download.file("https://data.somervillema.gov/download/gw4w-w7cw/application%2Fx-zip-compressed", temp)
unzip(temp, exdir = tdir)
parcels_data_source <- paste0(tdir, "/M274_parcels_CY23_FY24_sde.gdb")
st_layers(parcels_data_source)
```

First, let's try `M274TaxPar`; that's what this dataset *should* be about, though I know that the assessment data can be more fully featured.

```{r}
parcels <- st_read(parcels_data_source)
head(parcels)
```

Okay, it looks like this contains distinct buildings, which is good. What's not so good is that it doesn't tell us what material a building is made of.
Maybe the assessment data will tell that story:
```{r}
assess <- st_read(parcels_data_source, layer = "M274Assess")
gt(head(assess))
```

Jackpot.
Note the grain of this file: it's each building within each parcel.
This grain is important because it allows us to distinguish (detached) garages from the main buildings in tax parcels.

```{r}
assess %>%
filter(grepl("Garage", STYLE)) %>%
head() %>%
gt() %>%
tab_header(title = "FY24 Assessment Data")
```