---
title: "K+C Theme Demo"
author: Corey Runkel
date: December 16, 2022
format:
    revealjs:
        navigation-mode: vertical
#        logo: 
        footer: "Theme Demo"
---

# Problem

Technical demos in PowerPoint require lots of copying/pasting and tweaking for different audiences


# Idea

What if there were one presentation to rule them all?

. . .

What would that look like?


# Features

:::{.incremental}
- ~~hide~~ show the slides you want to show
- dive deeper for questions or digressions
- integrate images, diagrams, and plots
- show code well
- easy to view and export
:::


# Reveal.js + Quarto

Reveal
- JavaScript, with all the beauty of web design
- fine control over elements

Quarto
- ease of use
- integrated Python, R, SQL, shell
- nice documentation, IDE integration

# Integrated Code

Run code within presentation
```{{r}}
library(ggplot2)
```

. . .

but optionally hide it
```{r}
#| echo: fenced
#| include: false
library(ggplot2)
```

# Integrated plots
```{r}
ggplot(iris, aes(Petal.Length, Petal.Width)) +
    geom_point(aes(col = Species)) +
    labs(x = 'Length of petal', 'Width of petal')
```